---
title: "Getting Started with codiv"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with codiv}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# codiv: Host-Microbe Codiversification Analysis

Welcome to **codiv**! This package performs codiversification scans between pairs of phylogenetic trees to identify which branches show evidence of coevolution between hosts and their microbial symbionts.

---

## Quick Start

### In 30 seconds

The `codiv()` function takes three inputs and returns codiversification statistics:

```{r eval=FALSE}
library(codiv)

# Inputs:
# 1. Host phylogenetic tree (phylo object)
# 2. Symbiont phylogenetic tree (phylo object)
# 3. Host-to-symbiont linking table (data frame)

codiv_results <- codiv(
  Host_tree = Host_Tree,
  Symbiont_tree = Symbiont_Tree,
  Host_to_Symbiont_df = Host_to_Symbiont,
  permutations = 99
)

# Output: one row per scanned symbiont node, with statistics and p-values
head(codiv_results)
```

### What you get

`codiv()` returns a `codiv` object, which is a data frame with one row per scanned symbiont node plus `print` and `summary` methods for a quick overview:

```{r eval=FALSE}
codiv_results          # compact overview: settings and significant nodes per method
summary(codiv_results) # per-method statistic ranges and significant-node counts
```

The columns include:
- **Node identifiers** from the symbiont tree
- **Per-method test statistics** (Hommola's correlation, PACo goodness-of-fit, ParaFit global statistic) measuring how well symbiont topology matches host topology
- **Per-node permutation p-values**
- **Tree shape metrics** (Colless, Sackin indices)

Nodes with **high correlation + low p-value** are evidence of codiversification at that branch. To assess false discoveries for a whole scan, use `codiv_null_scans()`.

---

## Preparing Your Data

Before running `codiv()`, you need to prepare three pieces:

1. **Host phylogenetic tree**
2. **Symbiont phylogenetic tree**
3. **Metadata linking hosts to symbionts**

### 1. Phylogenetic Trees

Trees must be in R's `phylo` format (from the `ape` package). Examples:

```{r, eval=FALSE}
library(ape)

# Option A: Read from Newick/Nexus files
Host_Tree <- read.tree("host_phylogeny.nwk")
Symbiont_Tree <- read.tree("symbiont_phylogeny.treefile")

# Option B: Read from other formats
library(phangorn)
Host_Tree <- read.nexus("host_phylogeny.nex")

# Option C: Create from distance matrix
library(phangorn)
dist_matrix <- read.csv("distances.csv", row.names = 1)
Host_Tree <- upgma(as.dist(dist_matrix))
```

**Requirements:**
- Both trees must be binary (fully resolved)
- Tip labels must be unique within each tree
- Node labels are optional (codiv will create them if missing)

### 2. Host-to-Symbiont Linking Table

Create a data frame with exactly two columns: "Host" and "Symbiont"

```{r}
# Example structure
Host_to_Symbiont <- data.frame(
  Host = c("Mus_musculus", "Rattus_norvegicus", "Peromyscus_leucopus",
           "Mus_musculus", "Rattus_norvegicus"),
  Symbiont = c("MAG_0001__Bacteroides", "MAG_0002__Bacteroides",
               "MAG_0003__Bacteroides", "MAG_0004__Lactobacillus",
               "MAG_0005__Lactobacillus")
)

# Each row = one host-symbiont association
# Hosts are tips from the host tree
# Symbionts are tips from the symbiont tree
# For MAG data: each symbiont appears in only ONE host

head(Host_to_Symbiont)
```

**Tips:**
- For MAG/isolate data: each symbiont appears in exactly one host (one-to-one mapping)
- Column names MUST be exactly "Host" and "Symbiont"
- No missing values allowed
- Save as: CSV, TSV, Excel, or R object

### 3. Example Data Included with codiv

You don't need your own data to try codiv. The package ships example host and
symbiont trees and a linking table in `inst/extdata` — the rodent gut-microbiome
codiversification dataset from
[Sprockett et al. (2025)](https://www.nature.com/articles/s41467-025-57435-z),
21 rodent host species and several thousand bacterial MAGs — loaded with
`system.file()`:

```{r}
library(codiv)
library(ape)

Host_Tree <- read.tree(
  system.file("extdata", "Host_Tree.nwk", package = "codiv"))
Symbiont_Tree <- read.tree(
  system.file("extdata", "Symbiont_Tree.treefile", package = "codiv"))
Host_to_Symbiont <- read.csv(
  system.file("extdata", "Host_to_Symbiont.txt", package = "codiv"),
  sep = "\t")

# The bundled symbiont tree is unrooted, as most tree inference programs
# leave it. codiv() requires a rooted symbiont tree, because the scan tests
# one clade per internal node and the rooting decides which clades exist.
# Choose the rooting deliberately for real analyses (an outgroup is best);
# midpoint rooting is a reasonable default here.
Symbiont_Tree <- castor::root_at_midpoint(Symbiont_Tree)

# Inspect the data
Host_Tree
Symbiont_Tree
head(Host_to_Symbiont)
```

With `r ape::Ntip(Host_Tree)` hosts and several thousand symbionts, this is a
realistic scan. Running just the Hommola method and skipping the tree-distance
metrics keeps it to a few seconds:

```{r, message=FALSE, warning=FALSE}
codiv_results <- codiv(
  Host_tree = Host_Tree,
  Symbiont_tree = Symbiont_Tree,
  Host_to_Symbiont_df = Host_to_Symbiont,
  methods = "hommola",
  permutations = 99,
  subtree_features = FALSE,
  verbose = FALSE
)

codiv_results
```

codiv filters the symbiont tree's internal nodes (by span, tip count, and number
of hosts), scans the survivors, and returns one row per scanned node. Turn on the
other methods (`methods = c("hommola", "paco", "parafit")`) and
`subtree_features = TRUE` for a full analysis; those add runtime.

---

## Running codiv()

### Basic usage

```{r, eval=FALSE}
codiv_results <- codiv(
  Host_tree = Host_Tree,
  Symbiont_tree = Symbiont_Tree,
  Host_to_Symbiont_df = Host_to_Symbiont,
  permutations = 99,
  Save_fp = "codiv_results.tsv"
)
```

### Important parameters

| Parameter | Default | Meaning |
|-----------|---------|---------|
| `min_hosts` | 3 | Minimum number of different hosts in a symbiont subtree to include it in the scan (must be ≥ 3) |
| `min_symbiont_tips` | 7 | Minimum number of symbiont tips in a subtree (higher = more statistical power) |
| `span_fraction` | 0.1 | Only scan nodes with span ≤ 10% of max tree span (focuses on distal nodes) |
| `permutations` | 99 | Number of permutations for p-value calculation (higher = more precise) |
| `methods` | `c("hommola", "paco", "parafit")` | Codiversification methods (distance-based defaults), plus opt-in "topology" (branching-order test) |
| `cores` | all cores − 1 | Number of cores; nodes are scanned in parallel (single core on Windows) |
| `subtree_features` | TRUE | Calculate tree distance metrics? |
| `focus_hosts` | NULL | Optional host names; adds a `<host>_PRESENT` TRUE/FALSE column per host, for sorting the results by hosts of interest |
| `batch_size` / `n_batches` | auto | Control checkpoint frequency: nodes per batch, or number of batches (results are saved after each batch) |
| `verbose` | TRUE | Print progress messages? |

Symbiont tips with no host association are dropped automatically, and every node is scanned independently, so one problematic node cannot abort the run.

### Distance methods vs. the topology method

The three default methods (Hommola, PACo, ParaFit) are all **distance-based**: they correlate pairwise phylogenetic *distances*, so they are sensitive to branch lengths. Two trees with the same branching order but different evolutionary rates will score less than a perfect match.

The optional `"topology"` method asks a different question: **do the two trees have the same branching order?**, ignoring branch lengths entirely. It can flag congruence that the distance methods understate when rates differ between hosts and symbionts.

```{r, eval=FALSE}
codiv_results <- codiv(
  Host_tree = Host_Tree,
  Symbiont_tree = Symbiont_Tree,
  Host_to_Symbiont_df = Host_to_Symbiont,
  methods = c("hommola", "topology"),   # topology is opt-in
  permutations = 99
)
```

Because a tree-shape comparison needs one tip per host, the topology method reduces each host to its **largest single-host clade** (tie-broken by the tightest branch lengths). A host whose symbionts are completely scattered has no coherent position and is excluded. Two diagnostic columns make this visible:

- `Topology_hosts_used` — how many hosts contributed to the node's test
- `Topology_fraction_kept` — the fraction of symbiont tips retained after reducing each host to its dominant clade

A low `Topology_fraction_kept` means the congruence rests on heavy pruning; treat those nodes with caution, and note that nodes with fewer than three usable hosts return `NA`.

### Resuming interrupted runs

If your analysis stops before completing, resume it:

```{r, eval=FALSE}
# Continue from where it stopped
codiv_results <- codiv(
  Host_tree = Host_Tree,
  Symbiont_tree = Symbiont_Tree,
  Host_to_Symbiont_df = Host_to_Symbiont,
  Save_fp = "codiv_results.tsv",  # Same filename
  continue = TRUE,                 # Pick up where we left off
  permutations = 99
)
```

---

## Interpreting Results

### Output structure

```{r, eval=FALSE}
# Compact overview and per-method summary
codiv_results
summary(codiv_results)

# The full table (a data frame) for filtering / export
head(as.data.frame(codiv_results))

# Column meanings:
# Node_ID              - Symbiont tree node identifier
# N_Symbionts          - Number of tips in this subtree
# N_Hosts              - Number of different hosts in this subtree
# Symbiont_Colless     - Tree imbalance metric (higher = more imbalanced)
# Symbiont_Sackin      - Tree shape metric (related to tip distances)
# Hommola_r            - Correlation coefficient (-1 to 1)
# Hommola_pvalue       - Permutation p-value
# PACo_ss              - PACo goodness-of-fit sum of squares (if method = "paco")
# PACo_pvalue          - PACo permutation p-value (if method = "paco")
# ParaFitGlobal        - ParaFit global test statistic (if method = "parafit")
# ParaFit_pvalue       - ParaFit permutation p-value (if method = "parafit")
```

The object is still a data frame, so `$`, `[`, and `dplyr` verbs all work; use `as.data.frame(codiv_results)` when you want a plain data frame.

### Collapsing within-host variation

By default, every method runs on a **collapsed** subtree. Before the tests run, codiv collapses each monophyletic clade of symbionts from the **same host species** down to a single representative tip. This prevents inflated correlations that would otherwise arise when near-identical MAGs — for example genomes assembled from different individuals of the same host species — are counted as independent observations. If MAGs from one host species are **not** monophyletic (they fall in separate parts of the symbiont tree, reflecting genuinely distinct lineages), all of them are retained.

`Hommola_r` / `Hommola_pvalue`, `PACo_ss` / `PACo_pvalue`, and `ParaFitGlobal` / `ParaFit_pvalue` are all computed on this collapsed subtree. Setting `uncollapsed_hommola = TRUE` additionally runs Hommola's test on the *uncollapsed* subtree and adds `Uncollapsed_Hommola_r` / `Uncollapsed_Hommola_pvalue`; this is opt-in because it treats deeply sampled hosts as pseudoreplicated and is the most expensive statistic on large clades.

### Identifying codiversified nodes

Following [Sanders et al. (2023)](https://www.nature.com/articles/s41564-023-01388-w) and [Sprockett et al. (2025)](https://www.nature.com/articles/s41467-025-57435-z), a clade is called co-diversifying when it clears both a statistic and a p-value threshold, e.g. `Hommola_r > 0.75` and `Hommola_pvalue < 0.01`:

```{r, eval=FALSE}
library(dplyr)

significant_nodes <- as.data.frame(codiv_results) %>%
  filter(Hommola_r > 0.75,             # strong topological association
         Hommola_pvalue < 0.01) %>%
  arrange(desc(Hommola_r))

head(significant_nodes)
```

### Scan-wide false-discovery control

To gauge false discoveries for a whole scan, codiv implements a **second-order permutation test**: permute the host tree's tip labels, rerun the entire scan, and build a null distribution for the *number* of significant clades. This is the appropriate approach because nested clades in a tree are not independent, so a per-node multiple-testing correction (e.g. Benjamini-Hochberg) would be invalid here. `codiv_null_scans()` does this:

```{r, eval=FALSE}
null <- codiv_null_scans(
  Host_tree = Host_Tree,
  Symbiont_tree = Symbiont_Tree,
  Host_to_Symbiont_df = Host_to_Symbiont,
  n_permutations = 100,          # host-label permutations (outer null)
  stat_threshold = 0.75,         # r > 0.75 ...
  p_threshold = 0.01,            # ... and p < 0.01 define "co-diversifying"
  permutations = 999, methods = "hommola", cores = 8
)

null$observed        # significant clades in the real data
mean(null$null_counts)  # expected under permuted host labels
null$global_pvalue   # is the scan more co-diversifying than chance?
null$empirical_fdr   # expected false-discovery proportion at these thresholds
```

Because each permutation reruns the full scan, this captures exactly the dependence structure among nested clades. It is the recommended way to report significance for a scan.

### Tree shape metrics

The Colless and Sackin indices describe tree balance:
- **Colless index:** Higher values = more imbalanced tree (bifurcations uneven)
- **Sackin index:** Higher values = longer average path to tips

**Why it matters:** Different tree shapes can affect codiversification patterns. A perfectly balanced host tree with a highly imbalanced symbiont subtree might show weak correlation for reasons unrelated to coevolution.

---

## Secondary Analyses

The codiv package provides functions for deeper investigation of codiversification patterns:

### Leave-One-Host-Out Analysis

Assess which hosts drive the codiversification signal:

```{r, eval=FALSE}
loo_results <- loo_host_analysis(
  Host_tree = Host_Tree,
  Symbiont_tree = Symbiont_Tree,
  Host_to_Symbiont_df = Host_to_Symbiont,
  permutations = 99
)

# View importance scores
loo_results$host_importance

# Interpret: Hosts with high impact_score contribute disproportionately
# to the codiversification signal
```

**Faster option:** Test specific hosts only:

```{r, eval=FALSE}
# Just test 3 hosts (much faster)
loo_results <- loo_host_analysis(
  Host_tree = Host_Tree,
  Symbiont_tree = Symbiont_Tree,
  Host_to_Symbiont_df = Host_to_Symbiont,
  hosts_to_test = c("host_A", "host_B", "host_C"),
  permutations = 99
)
```

### Parameter Sensitivity Analysis

How robust are your results to parameter choices?

```{r, eval=FALSE}
sensitivity <- sensitivity_analysis(
  Host_tree = Host_Tree,
  Symbiont_tree = Symbiont_Tree,
  Host_to_Symbiont_df = Host_to_Symbiont,
  span_fraction_range = c(0.05, 0.1, 0.2),
  min_symbiont_tips_range = c(5, 7, 10),
  permutations = 99
)

# View sensitivity across parameters
sensitivity$sensitivity_data

# Strong codiversification should be robust across reasonable parameter ranges
```

### Visualizing Host-Symbiont Trees

Draw a tanglegram: the two trees face to face with a line for each association. This gives an intuitive sense of how congruent the trees are. Requires the `ggplot2` package and returns a ggplot object you can customize or save with `ggsave()`.

```{r, eval=FALSE}
# Lines colored by host (default)
plot_codiv_trees(
  Host_tree = Host_Tree,
  Symbiont_tree = Symbiont_Tree,
  Host_to_Symbiont_df = Host_to_Symbiont
)

# Highlight symbionts in significant nodes (needs a codiv result)
plot_codiv_trees(
  Host_tree = Host_Tree,
  Symbiont_tree = Symbiont_Tree,
  Host_to_Symbiont_df = Host_to_Symbiont,
  codiv_results = codiv_results,
  color_by = "significance",
  significance_threshold = 0.05,
  title = "Codiversification Patterns"
)
```

By default a cladogram layout is used (tips aligned, nodes spread by topology) and the symbiont tree is rotated to reduce line crossings. Set `use_branch_lengths = TRUE` to position nodes by branch length, or `show_symbiont_labels = TRUE` to label symbiont tips.

---

## Getting Help

### Troubleshooting

**"Tree must be binary"**
```{r, eval=FALSE}
# Solution: Collapse polytomies
library(ape)
Host_Tree <- multi2di(Host_Tree)
Symbiont_Tree <- multi2di(Symbiont_Tree)
```

**"Symbiont not found in tree"**
```{r, eval=FALSE}
# Check spelling and exact names
head(Symbiont_Tree$tip.label)    # Actual names in tree
head(Host_to_Symbiont$Symbiont)  # Names in linking table
# They must match exactly!
```

**Analysis running very slowly**

Runtime scales with the number of nodes scanned and permutations. The biggest
levers, roughly in order of impact:

```{r, eval=FALSE}
codiv(..., subtree_features = FALSE)   # skip the tree-distance metrics (often the
                                       #   most expensive step)
codiv(..., methods = "hommola")        # run one method instead of all three
codiv(..., permutations = 9)           # fewer permutations while exploring
codiv(..., min_symbiont_tips = 10,     # scan fewer, larger nodes ...
      span_fraction = 0.05)            #   ... focused on distal clades
codiv(..., cores = 8)                  # more cores (nodes scan in parallel;
                                       #   single core on Windows)
```

For very large trees (thousands of symbiont tips) a full run of all three methods
with `subtree_features = TRUE` can take hours and several GB of RAM; consider an
HPC node, or start with the hommola-only, metrics-off configuration above.

**Repeated `Found more than one class "phylo" in cache` messages**

This harmless message comes from R's `methods` package when two of codiv's
dependencies both register an S4 `phylo` class. It does not affect results. To
silence it for good, add this to your `~/.Rprofile`:

```{r, eval=FALSE}
.First <- function() {
  globalCallingHandlers(message = function(m) {
    txt <- conditionMessage(m)
    if (grepl("more than one class .+ in cache", txt) ||
        startsWith(txt, "Also defined by")) invokeRestart("muffleMessage")
  })
}
```

### Documentation & References

- **Function help:** `?codiv`, `?loo_host_analysis`, `?plot_codiv_trees`
- **Package documentation:** https://sprockettlab.github.io/codiv/
- **Methods papers:**
  - Hommola et al. 2009 (Biology Letters)
  - Balbuena et al. 2013 (Journal of Biogeography)

### Reporting Issues

Found a bug? Have a feature request?

Visit: https://github.com/SprockettLab/codiv/issues

Please include:
- Minimal reproducible example
- Error message (full traceback)
- Session info: `sessionInfo()`

---

## Session Information

```{r}
sessionInfo()
```
