---
title: "Secondary analyses"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 6, out.width = "100%")
```

Beyond the main scan, `codiv` provides functions for probing robustness, dating
co-diversification events, and visualizing host–symbiont congruence. All build on
a `codiv()` result, so they share its inputs and conventions.

```{r}
library(codiv)
sim <- simulate_codiv_data(n_hosts = 10, n_clades = 3, seed = 1)
```

## Which hosts drive the signal? Leave-one-host-out

`loo_host_analysis()` reruns the scan with each host removed in turn and reports
how much each host changes the number of significant nodes — a large
`impact_score` means the host contributes disproportionately to the signal.

```{r}
loo <- loo_host_analysis(sim$host_tree, sim$symbiont_tree, sim$links,
                         permutations = 99, methods = "hommola", verbose = FALSE)
loo$host_importance
```

## How sensitive are the results? Parameter sweeps

`sensitivity_analysis()` reruns the scan across a grid of `span_fraction` and
`min_symbiont_tips` values so you can see whether findings are stable or
threshold-dependent.

```{r}
sens <- sensitivity_analysis(sim$host_tree, sim$symbiont_tree, sim$links,
                             span_fraction_range = c(0.3, 0.6),
                             min_symbiont_tips_range = c(7, 10),
                             permutations = 99, methods = "hommola",
                             verbose = FALSE)
sens$sensitivity_data
```

## Dating co-diversification: the molecular clock

For strongly co-diversifying clades, `molecular_clock()` regresses symbiont
divergence against host clade age. A positive slope corroborates that symbiont
clades diversified alongside their hosts, and the slope estimates the symbiont
molecular rate (following
[Sanders et al. 2023](https://www.nature.com/articles/s41564-023-01388-w)).

```{r}
clock_sim <- simulate_codiv_data(n_hosts = 14, clades = list(
  list(codiversifying = TRUE, congruence = 1, n_per_host = 3)), seed = 1)
codiv_results <- codiv(clock_sim$host_tree, clock_sim$symbiont_tree,
                       clock_sim$links, span_fraction = 0.9, methods = "hommola",
                       permutations = 199, verbose = FALSE)

# the simulated host tree is ultrametric, so it stands in for a time-calibrated tree
mc <- molecular_clock(codiv_results, clock_sim$host_tree, stat_threshold = 0.75)
mc$slope
mc$r_squared
head(mc$per_clade)
```

## Visualizing congruence: tanglegrams

`plot_codiv_trees()` draws the host and symbiont trees face to face with a line
for each association — an intuitive read on how congruent they are. It returns a
`ggplot` object you can style or save with `ggsave()`.

```{r tanglegram}
plot_codiv_trees(sim$host_tree, sim$symbiont_tree, sim$links)
```

By default a cladogram layout is used and the symbiont tree is rotated to reduce
line crossings; set `use_branch_lengths = TRUE` to position nodes by branch
length, or pass a `codiv_results` object with `color_by = "significance"` to
highlight symbionts in significant clades.
