---
title: "Simulation & validation"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
```

Because real datasets never come with a ground truth, `codiv` includes
`simulate_codiv_data()` to build host/symbiont datasets where you *know* which
clades co-diversified. This is the basis for checking that the scan recovers
true signal and for calibrating false discovery with `codiv_null_scans()`.

## Simulating data with a known truth

```{r}
library(codiv)

sim <- simulate_codiv_data(n_hosts = 10, n_clades = 3, seed = 1)
names(sim)
```

`sim` contains a `host_tree`, a `symbiont_tree`, a `links` table, and a `truth`
table recording, for every simulated symbiont, which clade it belongs to and
whether that clade was generated to co-diversify:

```{r}
# one row per clade: is it co-diversifying, and how congruent with the host tree?
truth_by_clade <- unique(sim$truth[, c("clade", "codiversifying", "congruence")])
truth_by_clade
```

You control the design directly — number of hosts and clades, the proportion of
clades that co-diversify, how congruent they are, how many symbionts per host,
and the amount of branch-length noise. Passing an explicit `clades` list gives
per-clade control:

```{r eval=FALSE}
sim <- simulate_codiv_data(
  n_hosts = 12,
  clades = list(
    list(codiversifying = TRUE,  congruence = 1.0, n_per_host = 2),
    list(codiversifying = TRUE,  congruence = 0.7, n_per_host = 1),
    list(codiversifying = FALSE)                       # random association
  ),
  seed = 42
)
```

## Does the scan recover the signal?

```{r}
codiv_results <- codiv(sim$host_tree, sim$symbiont_tree, sim$links,
                       methods = "hommola", permutations = 99, verbose = FALSE)
summary(codiv_results)
```

Nodes drawn from co-diversifying clades should show high `Hommola_r` and low
p-values, while nodes from the randomly-associated clade should not.

## Scan-wide false-discovery control

A whole-scan question — *are there more significant clades than chance would
produce?* — cannot be answered by correcting per-node p-values, because nested
clades are not independent. `codiv_null_scans()` instead permutes the host tip
labels, reruns the entire scan, and builds a null distribution for the *number*
of significant clades:

```{r}
null <- codiv_null_scans(
  sim$host_tree, sim$symbiont_tree, sim$links,
  n_permutations = 20,             # host-label permutations (outer null)
  stat_threshold = 0.75,           # r > 0.75 ...
  p_threshold    = 0.01,           # ... and p < 0.01 define "co-diversifying"
  min_symbiont_tips = 7, span_fraction = 0.6,
  permutations = 99, methods = "hommola",
  verbose = FALSE
)

null$observed         # significant clades in the real data
mean(null$null_counts) # expected under permuted host labels
null$global_pvalue    # is the scan more co-diversifying than chance?
null$empirical_fdr    # expected false-discovery proportion at these thresholds
```

Because each permutation reruns the full pipeline, the null captures exactly the
dependence structure that a per-node correction ignores. This is the recommended
way to report significance for a scan (see also
[Sanders et al. 2023](https://www.nature.com/articles/s41564-023-01388-w)).

For serious calibration work, raise `n_permutations` (the outer null) and
`permutations` (the inner per-node null) and spread the work across `cores`.
